---
title: "vegan workshop - Eco-Data-Science"
author: 'Ann Bui, adapted by P. Spector'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

More details regarding ordination and visualization can be found [here](http://rpubs.com/an-bui/553683).

```{r libraries-data, message = FALSE}
# libraries
install.packages("tidyverse")
install.packages("vegan")
library(tidyverse)
library(vegan)
library(gdata)

setwd("/Users/PikesStuff/Desktop/Manuscript/Figures_&_Code/")
# data
# algal  communities
algae <- read.xls("Diversity.xls") %>% 
  column_to_rownames("Site")

# environmental variables
meta <- read.xls("Diversity_meta.xls")

# set up a "metadata" frame - will be useful for plotting later!
site_type <- meta %>% 
  # selects the named columns
  select(Site, Location)
```

### How speciose are my communities?

```{r species-richness}
#vegant::specsnumber() - counts the number of species in each site
#output is a named vector
sppr <- specnumber(algae)

#
sppr_aov <- aov(sppr ~ Location, data = site_type)
summary(sppr_aov)
```

```{r species-richness-plot}
# converting a named vector into a dataframe
sppr_df <- sppr %>% 
  # joined with site_type (metadata for easier plotting)
  enframe() %>% 
  # 
  full_join(site_type, by = c("name" = "Site"))

sppr_plot <- ggplot(sppr_df, aes(x = Location, y = value, color = Location)) +
  geom_boxplot()
sppr_plot
```

### How diverse are my communities?

```{r shannon-diversity}
# vegan::diversity() --> type into help page (Shannon diversity is default, and most commmon)
shannondiv <- diversity(algae)
#creates a named vector, Shannon diverisity for sites 
# try simpson's diversity
#modifiation of argument that diveristy takes
simpsondiv <- diversity(algae, index = "simpson")
```

```{r shannon-diversity-plot}
shandiv_df <- shannondiv %>% 
  # converting into a dataframe
  enframe() %>% 
  # joining with metadata
  full_join(., site_type, by = c("name" = "Site")) %>% 
  # calculating mean shannon diversity for each landtype
  group_by(Location) %>% 
  summarize(mean = mean(value))

shandiv_plot <- ggplot(shandiv_df, aes(x = Location, y = mean, fill = Location)) +
  geom_col()
shandiv_plot
```

## All code below is from an excellen workshop given by [An Bui](http://rpubs.com/an-bui/553683)

### How different are my communities in species composition?

#### perMANOVA
```{r bird-permanova}
# permutational analaysis of variance: are the centroids of my communities different?
bird_perm <- adonis(birds ~ landtype, data = env)
bird_perm
```

#### PCA

```{r bird-PCA}
# principale components analysis
# vegan::rda() <-make sure to look up difference between constrained and unconstrained analysis
birdPCA <- rda(birds)
birdPCA
# 

summary(birdPCA)
# Exp PCI explained 23.0590% of the variance (Eigenvalue); cumulative proportion explained is additive (across rows from proportion explained; add Eigenvalues)
# 

pcabiplot <- biplot(birdPCA)
# calls PCA oridination plot in base R, not super informative
# str(pcabiplot) --> next line calls this
#extracts coordinates from outplot of biplot (following call; PCAscores)
#Reminder: PCA bends axis to be most imformative

# 
PCAscores <- as.data.frame(pcabiplot$sites) %>% 
  bind_cols(site_type, .)

PCAvect <- as.data.frame(pcabiplot$species)


PCA_plot <- ggplot(PCAscores) +
  geom_point(aes(x = PC1, y = PC2, color = landtype)) +
  geom_segment(data = PCAvect, aes(x = 0, y = 0, xend = PC1, yend = PC2), arrow = arrow(length = unit(0.2, "cm"))) +
  geom_text(data = PCAvect, aes(x = PC1, y = PC2, label = rownames(PCAvect)))

PCA_plot
```

#### NMDS

Non-metric Multidimensional Scaling

```{r bird-NMDS}
#vegan::metaMDS
bird_NMDS <- metaMDS(birds)

bird_NMDS

nmdsplot <- plot(bird_NMDS)
#plot in base R
#mapping of dissimilarirty; axis don't matter
#circles = sites, + = species
stressplot(bird_NMDS)
#stress is a metric of how far a community has moved (in terms of dissimilarity); spread of stress, each point is a community. 
```

```{r NMDS-plot}
# extract outputs of nMDS into data frame
# 
nmds_df <- as_tibble(bird_NMDS$points) %>% 
  # binding it with metadata
  bind_cols(site_type, .)

nmds_plot <- ggplot(nmds_df, aes(x = MDS1, y = MDS2, color = landtype, shape = landtype)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse()
nmds_plot

#?betadisper --> looks at dispersion from the centroid, useful in comparison to PERMANOVA 
```

##### Things to consider about stress

```{r subsampled-NMDS}
sub <- birds[sample(nrow(birds), 20), ]
subNMDS <- metaMDS(sub)
stressplot(subNMDS)
#reducing to 20 interations gives acceptable stress (for this dataset)
#nFit explores best fit in nMDS --> look in to this
```

### How is community structure related to specific environmental variables?

```{r bird-CCA}
# Canonical Correspondence Analysis
# only shows vartion in communities based on enironmental variables (4 columns in env_var)
# Hypothesis test for looking at which variables contribute to community compostion; do *not* run all variables at once; different than exploring in nMDS
birdCCA <- cca(birds ~ canopy_height + stems_ha + big_stem_bas, data = env) #tests these three strucutural variables
birdCCA
#constrained vs unconstrained inertia; there is variation that is *not* explained by these three variables alone (canopy_height + stems_ha + big_stem_bas)

```

```{r bird-CCA-plot}
ccaplot <- plot(birdCCA)
# similar to PCA; x and y axis are most explained variables in the community 

# extraxting coordaintes from the biplot
#scaling factor is taken out of the structure of the plot: str(ccaplot)
ccavectors <- as.data.frame(ccaplot$biplot * 5.15) #<-- last number changes based on what you see from str(ccaplot)

#coordinates for sites and species from biplot output
site_data <- as.data.frame(ccaplot$sites) %>% 
  bind_cols(site_type, .)

species_data <- as.data.frame(ccaplot$species)

cca_plot <- ggplot(site_data) +
  geom_point(aes(x = CCA1, y = CCA2, color = landtype), shape = 19) +
  geom_segment(data = ccavectors, aes(x = 0, y = 0, xend = CCA1, yend = CCA2), arrow = arrow(length = unit(0.2, "cm"))) +
    #scale_x_continuous(limits = c(-10, 16)) +
    #scale_y_continuous(limits = c(-3, 12)) +
  geom_point(data = species_data, aes(x = CCA1, y = CCA2), shape = 17, size = 2, color = "blue") +
  geom_text(data = ccavectors, aes(x = CCA1, y = CCA2, label = rownames(ccavectors)))
cca_plot
```